# Problem Set 1 {.unnumbered}

This is a preliminary version of problem set 1. The final version will be shared through GitHub.

Reminders:

* Add a title to all your graphs.
* Add a label to the x and y axes when not obvious what they are showing.
* Think about transformations that convey the message in clearer fashion.

## Measles 

(@) Load the __dslabs__ package and figure out what is in the `us_contagious_diseases` dataset. Create a data frame, call it `avg`, that has a column for `year`, and a `rate` column containing the cases of Measles per 10,000 people per year in the US. Because we start in 1928, exclude Alaska and Hawaii. Make sure to take into account the number of weeks reporting each year. If a week was not report, it should not be included in the calculation of the rate. 
```{r}
library(tidyverse)
library(dslabs)
avg <- us_contagious_diseases |>
  filter(!state %in% c("Hawaii","Alaska") & disease == "Measles" & weeks_reporting > 0) |>
  group_by(year) |>
  summarize(rate = sum(count*52/weeks_reporting, na.rm = TRUE)/sum(population) * 10000)
```

(@) Use the data frame `avg` to make a trend plot showing the cases rate for Measles per year. Add a vertical line showing the year the Measles vaccines was introduced. Write a short paragraph describing the graph to someone you are urging to take the Measles vaccines.
```{r}
library(ThemePark)
avg |> ggplot(aes(year, rate)) + geom_line() + 
  geom_vline(xintercept = 1963, color = "blue") + 
  theme_barbie()
```

(@) Is the pattern observed above the same for each state? Add a grey trend line for each state to the plot above. Use a transformation that keeps the high rates from dominating the figure.
```{r}
us_contagious_diseases |>
  filter(disease == "Measles" & weeks_reporting > 0) |>
  mutate(rate = count*52/weeks_reporting/population * 10000) |>
  ggplot(aes(x=year)) + 
  geom_line(aes(y = rate, group = state), color = "grey", alpha = 0.5) + 
  geom_line(date = avg, aes(x = year, y = rate)) +
  scale_y_continuous(trans = "sqrt") +
  geom_vline(xintercept = 1963, color = "blue")
```

(@) In the plot above we can't tell which state is which curve. Using color would be challenging as it is hard if not impossible to find 48 colors we humans can distinguish. To make a plot where you can compare states knowing which is which, use one of the axis for state and the other for year, and then use hue or intensity as a visual cue for rates. Use a `sqrt` transformation to avoid the higher rates taking up all the color scale. Use grey to denote missing data. Order the states based on their highest peak.
```{r}
library(RColorBrewer)
mycolors <- brewer.pal(9, "Reds")
us_contagious_diseases |>
  filter(!state %in% c("Hawaii","Alaska") & disease == "Measles") |>
  mutate(rate = count/population*10000*52/weeks_reporting,
         state = reorder(state, rate, max, na.rm = TRUE)) |>
  ggplot(aes(year, state, fill = rate)) +
  geom_tile(color = "grey") +
  scale_x_continuous(expand = c(0,0)) +
  scale_fill_gradientn(colors = mycolors, trans = "sqrt") +
  geom_vline(xintercept = 1963, color = "blue") +
  theme_minimal() +  
  theme(panel.grid = element_blank(), 
        legend.position = "bottom", 
        text = element_text(size = 8)) +
  labs(title = "Measles cases per year in the US", x = "", y = "")
```


## COVID-19


(@) The csv file shared [here](https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-data.csv) includes weekly data on SARS-CoV-2 reported cases, tests, COVID-19 hospitalizations and deaths, and vaccination rates by state. 
 * Import the file into R without making a copy on your computer. 
 * Examine the dataset.
 * Write a sentence describing each variable in the dataset.
```{r}
url <- "https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-data.csv"
dat <- read_csv(url)
```
state: This variable contains the abbreviation for the U.S. state.

state_name: This variable contains the full name of the U.S. state corresponding to the state abbreviation.

population: This variable represents the population of the state.

region: This variable appears to represent a region code or category.

mmwr_year: This variable represents the year of data collection.

mmwr_week: This variable represents the week of the year for data collection.

cases: This variable represents the number of reported SARS-CoV-2 cases for the specific week and state.

tests: This variable represents the total number of COVID-19 tests conducted in the state during the week.

hosp: This variable represents the count of COVID-19 hospitalizations in the state during the week.

deaths_underlying_cause: This variable represents the number of COVID-19 related deaths where COVID-19 was the underlying cause.

deaths_multiple_causes: This variable appears to represent the number of COVID-19 related deaths where COVID-19 was one of multiple contributing causes.

deaths_prov: This variable may represent provisional or reported COVID-19 deaths.

series_complete: This variable may relate to the number of individuals who have completed a series of COVID-19 vaccinations.

booster: This variable may represent the number of individuals who have received a COVID-19 booster dose.

bivalent: This variable may be related to the number of individuals who have received a bivalent COVID-19 vaccine.


(@) One of these columns could benefit from being a factor. Identify the column and convert it to factor.
```{r}
dat <- mutate(dat, region = factor(region))
```

(@) Rather than providing a date, the dataset provides MMWR year and week. Read [this document](https://ndc.services.cdc.gov/wp-content/uploads/MMWR_Week_overview.pdf) and write a function to convert these to the start of the MMWR week in ISO-8601.
```{r}
library("lubridate")
mmwr_to_date <- function(mmwr_year, mmwr_week){
  first_day <- floor_date(make_date(mmwr_year, 1, 4) , unit = "week")
  date <- first_day + weeks(mmwr_week - 1)
  return(date)
}
```

(@) Add a columns `start_date` and `end_date` with the start and end of the MMWR week. Confirm that it worked by computing the MMWR week and year for both start and end date and comparing it to the MMWR week and year provided.
```{r}
dat <- dat |> mutate(start_date = mmwr_to_date(mmwr_year, mmwr_week),
                     end_date = start_date + days(6))
```

```{r}
dat <- dat |> mutate(start_date = mmwr_to_date(mmwr_year, mmwr_week),
                   end_date = start_date + days(6))
## check : these should all be TRUE
dat |> summarize(w1 = all(epiweek(start_date) == mmwr_week),
                 y1 = all(epiyear(start_date) == mmwr_year),
                 w2 = all(epiweek(end_date) == mmwr_week),
                 y2 = all(epiyear(end_date) == mmwr_year))
```

(@) Make a trend plot similar to the one we made for Measles:

* Include a trend line for the US cases rate. Use per 100,000 person per week as the unit.
* Add a trend for each state, with color representing region.
* Use the end of the week date for your x-axis.
* Add a vertical dashed line on the day COVID-19 vaccination started being delivered in the US.

Write a paragraph describing the COVID-19 pandemic by describing the plot.
```{r}
library(tidyverse)
library(lubridate)
library(RColorBrewer)

# Define the URL for the COVID-19 dataset
url <- "https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-data.csv"

# Read the COVID-19 dataset
covid_data <- read_csv(url)

# Convert MMWR year and week to date
covid_data <- covid_data %>%
  mutate(start_date = mmwr_to_date(mmwr_year, mmwr_week),
         end_date = start_date + days(6))

# Filter out rows with missing data
covid_data <- covid_data %>%
  filter(!is.na(cases) & !is.na(population) & !is.na(region))

# Calculate the COVID-19 case rate per 100,000 people per week
covid_data <- covid_data %>%
  mutate(rate = cases / population * 100000)

# Create the trend plot with a predefined color palette
covid_plot <- covid_data %>%
  ggplot(aes(x = end_date)) +
  geom_line(aes(y = rate, group = state, color = as.factor(region)), alpha = 0.5) +
  geom_line(data = covid_data %>%
              group_by(end_date) %>%
              summarize(avg_rate = sum(rate, na.rm = TRUE) / n()),
            aes(y = avg_rate), color = "blue") +
  geom_vline(xintercept = ymd("2020-12-14"), linetype = "dashed", color = "red") +
  scale_color_brewer(palette = "Set1") +  # Use the Set1 color palette
  labs(title = "COVID-19 Cases Rate in the US by State",
       x = "End of MMWR Week",
       y = "Cases Rate per 100,000 People per Week") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Print the plot
print(covid_plot)
```



(@) The number of cases depends on testing capacity. Note that during the start of the pandemic, when we know many people died, there are not that many cases reported. Also notice somewhat large variability across states that might not be consistent with actual prevalence. The `tests` columns provides the cumulative number of tests performed by the data represented by the week. This data is not official CDC data. It was provided by Johns Hopkins Coronavirus Resource Center. Before using the data, explore the data for each state carefully looking for potential problems.

For each state compute and plot the number of tests perforemd each week. Look at the plot for each state and look for possible problems. No need to make this plot pretty since we are just using it for data exploration. Report any inconsistencies if any.
```{r}
dat |>
  filter(!is.na(tests)) |>
  arrange(end_date) |>
  group_by(state) |>
  mutate(tests = diff(c(0,tests))) |>
  ggplot(aes(end_date, tests/10^5)) + geom_line() +
  facet_wrap(~state, scale = "free_y")
```
The overall trend for each states is increasing. But, the rate of increasing is inconsistent among them. 

(@) To see if the inconsistencies seen in the previous plot are a problem if we are only going to look at the total number of tests at the end, plot the cumulative tests for each of the states with inconsistencies and see if the results are sensible. Explain your answer in 1-2 sentences.
```{r}
library(tidyverse)

# Filter out rows with missing test data and arrange by end_date
test_data <- dat %>%
  filter(!is.na(tests)) %>%
  arrange(end_date)

# Calculate cumulative tests for each state
cumulative_tests <- test_data %>%
  group_by(state) %>%
  mutate(cumulative_tests = cumsum(tests))

# Identify states with inconsistencies in the cumulative tests
inconsistent_states <- cumulative_tests %>%
  group_by(state) %>%
  summarize(max_diff = max(diff(cumulative_tests))) %>%
  filter(max_diff > 0) %>%
  pull(state)

# Filter data for states with inconsistencies and create cumulative test plots
cumulative_test_plots <- cumulative_tests %>%
  filter(state %in% inconsistent_states) %>%
  ggplot(aes(end_date, cumulative_tests / 10^5, color = state)) +
  geom_line() +
  labs(title = "Cumulative Tests Over Time",
       x = "End of MMWR Week",
       y = "Cumulative Tests (per 100,000)")

# Print the cumulative test plots for inconsistent states
print(cumulative_test_plots)
```
Basically, the cumulative testing data aligns with expectations. The inconsistencies observed in previous graph have a substantial impact on the overall cumulative testing figures.

(@) JHU stopped reporting some time in 2020. What was that date? Show the day in the format September 18, 2022.
```{r}
library(lubridate)

# Filter the data for Maryland (MD)
md_data <- dat %>%
  filter(state == "MD") %>%
  arrange(desc(end_date))

# Find the last available date for Maryland
last_reporting_date <- md_data %>%
  filter(!is.na(tests)) %>%
  slice(1) %>%
  pull(end_date)

# Format the last reporting date in the desired format
formatted_date <- format(last_reporting_date, "%B %d, %Y")

formatted_date
```

(@) Compute the number of tests per capita for the last day JHU reported these statistics. Make a boxplot of these values for each region and include the state level data with the state abbreviation as a label. Write a sentences describing these differences you see and how this could affect our interpretation of differences in cases rates across states. 
```{r}
last_day <- as_date("2021-1-2") ## this should come from the previous exercise... 
dat |> filter(end_date == last_day) |>
  ggplot(aes(region, tests/population)) +
  geom_boxplot() +
  geom_text(aes(label = state)) +
  labs(x = "Region", y = "Test per capita", title = "SARS-COV2 tests per person")
```
MA, CT, VT, ME, NH, which have wider ranges of testing rates and higher testing rates, have more varied and potentially skewed case rate data and may appear to have a higher case rate than other regions. 

(@) Although JHU stopped collecting testing data from the states, the CDC collected data from a few laboratories. We provide these date in [this url](https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-tests.txt).

* Import the data into R without downloading the file.
* Make sure that you create a data frame with a column with dates in `Dates` format and tests as numbers.
```{r}
url <- "https://raw.githubusercontent.com/datasciencelabs/2023/main/data/covid19-tests.txt"
tests <- read_delim(url, delim = " ")
```

```{r}
tests <- tests |> mutate(date = mdy(date), tests = parse_number(tests))
## alternatively we can re-read the file
tests <- read_delim(url, delim = " ", 
                    col_types = cols(date = col_date(format = "%m-%d-%y"),
                                     tests = col_number()))
```

(@) Plot the tests per week to see the trend in time. Write a sentence of the pattern you see.
```{r}
library(ggplot2)

# Create a plot of tests per week
ggplot(tests, aes(x = date, y = tests)) +
  geom_line() +
  labs(title = "Trend of COVID-19 Tests Conducted per Week",
       x = "Date",
       y = "Tests Conducted") +
  theme_minimal()
```

At the start of the pandemic there were few tests conducted and now we are also seeing a drop.

(@) The analysis on tests points to cases as not being a good measure of the state of the pandemic. Remake the trend plot but using death rates instead of cases rates. Write a sentence on how this plot better shows about the start of the pandemic that the cases plot did not. 
```{r}
# Calculate death rates per 100,000 people per week
dat <- dat |>
  filter(!is.na(deaths_prov)) |>
  mutate(death_rate = (deaths_prov / population) * 100000)

# Create a trend plot of death rates
ggplot(dat, aes(x = end_date, y = death_rate)) +
  geom_line() +
  labs(title = "Trend of COVID-19 Death Rates per 100,000 People per Week",
       x = "End of Week Date",
       y = "Death Rates per 100,000 People") +
  theme_minimal()
```

We see that the highest death rates were actually at the start of the pandemic and that is was particularly bad in the northeast.


(@) We want to examine the percent of the population that completed the first series of vaccines, received the booster, and received the bivalent booster, respectively. First run this line of code and examine what it does.

```{r}
#| eval: false
tmp <- dat |> 
  pivot_longer(c(series_complete, booster, bivalent), names_to = "series", values_to = "percent") |>
  select(state, region, population, end_date, series, percent) |>
  filter(!is.na(percent)) |>
  mutate(percent = percent/population,
         series = factor(series, c("series_complete", "booster", "bivalent"))) 
```

Then make a plot showing the percent of population vaccination for each state. Use color to represent region. 

* Show the dates on the x axis with the month abbreviation and year. 
* Place the three plots vertically, on top of each other. 
* Show percentages on the y axis. Hint: use `scales::percent`.
```{r}
library(ggplot2)
library(scales)

# Create the plot
ggplot(tmp, aes(x = end_date, y = percent, fill = region)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(series ~ ., scales = "free_y", space = "free_y") +
  scale_x_date(labels = date_format("%b %Y"), date_breaks = "1 month") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(title = "Vaccination Coverage by Series and State",
       x = "Date",
       y = "Percentage of Population Vaccinated") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

(@) For each period used to make the trend plot for the three series, make a boxplot showing the maximum percentage reached by every state stratified by region. Let the range of the y axis adapt to the range of each group.
```{r}
library(ggplot2)

# Create a new column for the month-year period
tmp <- tmp %>%
  mutate(period = format(end_date, "%b %Y"))

# Create boxplots
ggplot(tmp, aes(x = period, y = percent, fill = region)) +
  geom_boxplot() +
  facet_wrap(~series, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), expand = c(0, 0)) +
  labs(title = "Maximum Percentage Vaccinated by State and Region",
       x = "Month-Year Period",
       y = "Maximum Percentage Vaccinated") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

(@) Use the plot above to define four periods: No vaccine, First vaccine, Booster, and Bivalent. 
Get the dates when each vaccine series starts (these don't have to be exact). Create a version of your data frame with a column called period that keeps the period associated with the week contained in each row.
```{r}
# Define approximate start dates for each period based on the plot
no_vaccine_start <- as.Date("2020-01-01")  # Before vaccinations started
first_vaccine_start <- as.Date("2020-12-01")  # Start of the first vaccine series
booster_start <- as.Date("2021-08-01")  # Start of booster shots
bivalent_start <- as.Date("2022-02-01")  # Start of bivalent booster shots

# Create a new column 'period' based on the start dates
tmp <- tmp %>%
  mutate(period = case_when(
    end_date < first_vaccine_start ~ "No vaccine",
    end_date < booster_start ~ "First vaccine",
    end_date < bivalent_start ~ "Booster",
    TRUE ~ "Bivalent"
  ))
```

(@) Define a new variable that stores the maximum vaccination percentage reached during each period. But for the first vaccine period use `series_complete`, for the booster period use the `booster` column, and for the bivalent period use the bivalent percentages. Remove data from the no vaccine period. The make a plot comparing the COVID-19 death rate to the percent vaccinated. Use color to denote region. You should produce three plots, each with it's own axes range. Put the three plots in three entries of a 2x2 layout. Comment on what you observe.
```{r}
library(dplyr)
library(ggplot2)

# Filter out data from the "No vaccine" period
filtered_tmp <- tmp %>%
  filter(period != "No vaccine")

# Create a new variable for maximum vaccination percentage during each period
vaccination_max <- filtered_tmp %>%
  group_by(state, region, period) %>%
  summarize(max_percent_vaccinated = max(percent)) %>%
  pivot_wider(names_from = period, values_from = max_percent_vaccinated)

# Merge vaccination data with COVID-19 death rate data
merged_data <- dat %>%
  left_join(vaccination_max, by = c("state", "region"))

# Create three separate plots, each with its own axes range
plot1 <- ggplot(merged_data, aes(x = death_rate, y = `First vaccine`, color = region)) +
  geom_point() +
  labs(title = "COVID-19 Death Rate vs. First Vaccine Percentage",
       x = "COVID-19 Death Rate",
       y = "First Vaccine Percentage") +
  theme_minimal()

plot2 <- ggplot(merged_data, aes(x = death_rate, y = `Booster`, color = region)) +
  geom_point() +
  labs(title = "COVID-19 Death Rate vs. Booster Percentage",
       x = "COVID-19 Death Rate",
       y = "Booster Percentage") +
  theme_minimal()

plot3 <- ggplot(merged_data, aes(x = death_rate, y = `Bivalent`, color = region)) +
  geom_point() +
  labs(title = "COVID-19 Death Rate vs. Bivalent Percentage",
       x = "COVID-19 Death Rate",
       y = "Bivalent Percentage") +
  theme_minimal()

# Arrange the three plots in a 2x2 layout
library(gridExtra)
grid.arrange(plot1, plot2, plot3, ncol = 2)
```

(@) Does population density have an effect on infections? Use the `state.area` predefined variable to add compute population density. Make a histogram and density plot of state densities. Note that you will have to add the geographical area for Puerto Rico and DC as it is not included in `state.area`
```{r}
my.state.abb <- c(state.abb, "PR", "DC")
my.state.area <- c(state.area, 5325, 69)
popdens <- dat |> filter(end_date == min(end_date)) |> 
  select(state, population) |>
  mutate(area = my.state.area[match(state, my.state.abb)]) |> 
  mutate(popdens = population / area) 
  
popdens |> ggplot(aes(popdens)) +
  geom_histogram(aes(y = after_stat(density)), bins = 25, color = "black") +
  geom_density() +
  scale_x_log10() +
  labs(title = "Distribution of poplation density across states", x = "Population density", y = "Density")
```

(@) Plot death rates versus density for the four periods defined above.
```{r}
# Merge population density data with COVID-19 death rate data
merged_data <- dat %>%
  left_join(popdens, by = c("state"))

# Create separate plots for each period
plot_no_vaccine <- ggplot(merged_data, aes(x = popdens, y = death_rate, color = region)) +
  geom_point() +
  labs(title = "Death Rate vs. Population Density (No Vaccine Period)",
       x = "Population Density",
       y = "Death Rate") +
  theme_minimal()

plot_first_vaccine <- ggplot(merged_data, aes(x = popdens, y = death_rate, color = region)) +
  geom_point() +
  labs(title = "Death Rate vs. Population Density (First Vaccine Period)",
       x = "Population Density",
       y = "Death Rate") +
  theme_minimal()

plot_booster <- ggplot(merged_data, aes(x = popdens, y = death_rate, color = region)) +
  geom_point() +
  labs(title = "Death Rate vs. Population Density (Booster Period)",
       x = "Population Density",
       y = "Death Rate") +
  theme_minimal()

plot_bivalent <- ggplot(merged_data, aes(x = popdens, y = death_rate, color = region)) +
  geom_point() +
  labs(title = "Death Rate vs. Population Density (Bivalent Period)",
       x = "Population Density",
       y = "Death Rate") +
  theme_minimal()

# Arrange the four plots in a grid layout
library(gridExtra)
grid.arrange(plot_no_vaccine, plot_first_vaccine, plot_booster, plot_bivalent, ncol = 2)
```

